Accelerated convergence of convective simulations using boundary value problems

### 2017-12-29 tex/bvp_paper.tex
adds to-do list to paper
first draft of figure captions and discussion

@@

\section{Experiment}
\label{sec:experiment}
In our study, we adopt the Oberbeck-Boussinesq approximation.  Here, the
fluid has constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$).  We non-dimensionalize length by the layer height ($L_z$) ,
temperature by the (constant) initial temperature gradient across the layer ($\grad T_0$), and time
by the freefall timescale ($L_z / v_{\text{ff}}$, with $v_{\text{ff}} = \sqrt{\alpha g L_z^2 \grad T_0}$, where $g$ is 
uniform gravitational acceleration in the $-\hat{z}$ direction). The dimensionless Boussinesq
equations governing the velocity $\bm{u} = u\hat{x} + v\hat{y} + w\hat{z}$, temperature
$T = T_0 + T_1$, and reduced pressure $\varpi$ are \cite{spiegel&veronis1960}
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy}
\end{gather}
where the dimensionless control parameters are the Rayleigh and Prandtl numbers,
\begin{equation}
\text{Ra} = \frac{g \alpha L_z^4 \grad T_0}{\nu\chi} = \frac{(L_z\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}.
\end{equation}
The dimensionless vertical extent of the domain is $z = [-1/2, 1/2]$, and at the boundaries
we impose no-slip, impenetrable boundary conditions such that $w = u = v = 0$ at $z = \pm 1/2$.
At the lower boundary, we employ a fixed flux condition such that $\partial T_1 / \partial z = 0$
at $z = -1/2$, and we impose a fixed temperature condition $T_1 = 0$ at $z = 1/2$. Both
horizontal directions are periodic, extending over a range $x, y = [0, \Gamma]$, where
the aspect ratio is $\Gamma = 2 \lambda_c$, with
$\lambda_c = 2\pi/k_c$, and $k_c=2.5519$ is the wavenumber of convective onset for this choice of
boundary conditions.

@@

\subsection{The Boundary Value Problem}
\begin{figure}[b]
\includegraphics[width=\textwidth]{./figs/time_trace.png}
\caption{\label{fig:time_trace} }
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/pdf_comparison.png}
\caption{\label{fig:pdf_comparison} }
\end{figure}

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{\label{fig:temp_comparison} }
\end{figure}

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/parameter_space_comparison.png}
\caption{\label{fig:parameter_space_comparison} }
\end{figure}

\section{Discussion \& Conclusions}
\label{sec:results}
I don't have the mental stamina to write this section right now.  In general, here's what it's going to say:
\begin{enumerate}
\item This procedure works, as we've shown.
\item How to apply this to more complicated cases, like stratified convection
\item How to apply this to different boundary conditions
\item Brief discussion of other methods that can be used to fast-forward, e.g., bootstrapping.
\item This is just a first step.
\end{enumerate}

>>>>

\section{To Do}
\begin{enumerate}
\item Get 3D cases in
\item Continue Improving Figures
\item Write figure captions
\item Think about post-bvp multiplying.  Should it be a profile for both the velocity
and for the T perturbations, rather than a scalar?  I'm starting to think it should.
\end{enumerate}

@@

\section{Experiment}
\label{sec:experiment}
In our study, we adopt the Oberbeck-Boussinesq approximation.  Here, the
fluid has constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$).  We non-dimensionalize length by the layer height ($L_z$) ,
temperature by the (constant) initial temperature gradient across the layer ($\grad T_0$), and time
by the freefall timescale ($L_z / v_{\text{ff}}$, with $v_{\text{ff}} = \sqrt{\alpha g L_z^2 \grad T_0}$, where $g$ is 
uniform gravitational acceleration in the $-\hat{z}$ direction). The dimensionless Boussinesq
equations governing the velocity $\bm{u} = u\hat{x} + v\hat{y} + w\hat{z}$, temperature
$T = T_0 + T_1$, and reduced pressure $\varpi$ are \cite{spiegel&veronis1960}
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy}
\end{gather}
where the dimensionless control parameters are the Rayleigh and Prandtl numbers,
\begin{equation}
\text{Ra} = \frac{g \alpha L_z^4 \grad T_0}{\nu\chi} = \frac{(L_z\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}.
\end{equation}
The dimensionless vertical extent of the domain is $z = [-1/2, 1/2]$, and at the boundaries
we impose no-slip, impenetrable boundary conditions such that $w = u = v = 0$ at $z = \pm 1/2$.
At the lower boundary, we employ a fixed flux condition such that $\partial T_1 / \partial z = 0$
at $z = -1/2$, and we impose a fixed temperature condition $T_1 = 0$ at $z = 1/2$. Both
horizontal directions are periodic, extending over a range $x, y = [0, \Gamma]$, where
the aspect ratio is $\Gamma = 2$.

@@

\subsection{The Boundary Value Problem}
\begin{figure}[b]
\includegraphics[width=\textwidth]{./figs/time_trace.png}
\caption{Traces of system energies vs. time for a long thermal rundown (a) and BVP convergence
(b) are shown for Ra = $1.30 \cdot 10^7$ ($S = 10^4$).  The two plots are scaled such that
one time unit on the x-axis of either run takes up an equivalent amount of space on the paper
to illustrate the time savings of the BVP run.  (c\&d) Flux plots, where black is the sum of the
flux, blue is enthalphy flux and red is condcutive flux. (c) Fluxes are shown for the IVP at an early
time to illustrate the asymmetry of the flux through the system. (d) Fluxes through the converged,
rundown state are compared to the fluxes in the converged BVP solution, where their difference
is shown in (e). \label{fig:time_trace} }
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/pdf_comparison.png}
\caption{We compare the time-variant dynamics in our 2D solutions by showing probability distribution
functions of the values of vertical velocity (a), horizontal velocity (b), and nonlinear heat
transport (c) in our converged solutions.  The horizontal velocities are very similar, with a
KS test similarity of X.  The vertical velocities, too, are similar, with a KS similarity of
Y.  The nonlinear heat transport, while overall quite similar, shows that the BVP experiences
a greater number of extreme events.  The peak around zero for all three PDFs reflects the
no-slip boundary conditions.\label{fig:pdf_comparison} }
\end{figure}

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{Here we compare the evolved thermodynamic state of a BVP to that of a rundown IVP that
has been run through a thermal timescale.  (a) The temperature profiles, as a function of height,
are shown. (b) The percentage difference between the temperature profiles, as a function of height,
is shown.  (c) PDFs comparing the values of the point-by-point temperatures over the averaging
window.  While the median of the PDF has a distinct offset between the two runs, the
thermal fluctuations about the mean which drive the convection are nearly the same.
\label{fig:temp_comparison} }
\end{figure}

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/parameter_space_comparison.png}
\caption{Here we demonstrate the efficacy of the BVP method at many values of Ra.  We show
(a) Nusselt number scaling, (b) Reynolds number scaling, and (c) scaling of the mean temperature
over the whole simulation.  (a) The Nusselt number follows a classic 2/7 scaling law
\cite{johnston&doering2009}, until it reaches S = $10^{3 + 2/3}$, at which point the
two-roll solution begins oscillating horizontally, showing time-variant Nu.  The range of Nu
over the averaging window is (blah), and the trend of Nu (the variance of the mean Nu) is shown
as (blah).  (b) Re follows an expected 1/2 scaling law. (c) The mean value of temperature, minus
the value at the fixed-temperature boundary, is shown.  As Ra increases, the boundary layers shrink,
and the convection becomes more efficient, we anticipate this value to get closer and closer to
zero as the atmosphere approaches the isotherm defined by the upper boundary.
\label{fig:parameter_space_comparison} }
\end{figure}

\section{Discussion \& Conclusions}
\label{sec:results}
While imperfect, the method presented here is a first step towards taking meaningful measurements
of highly turbulent convection on manageable, human timescales.  As demonstrated in Figs.
1-4, this BVP method quickly converges simulations to within a few percent of the true final
state, while preserving the natural behavior of the convective solution (such as the oscillatory
nature of high-Ra 2D states in \ref{fig:parameter_space_comparison}).

While not perfect, the BVP method here has one major benefit over some of the other methods
currently being used to achieve high Ra solutions.  One oft-used method is that of bootstrapping,
in which the converged solution of a low-Ra state is used as initial conditions for a higher-Ra
simulation.  While these methods are extremely powerful, they are influenced by hysteresis effects,
and the powerful, steady rolls achieved at low Ra can result in an artifically over-stable high-Ra
roll solution.  The BVP method can be used with random noise initial conditions which allow the
convective solution to be naturally chosen by the dynamics.

One benefit of the method presented here is that it can be easily extended to more complicated
configurations.  For example, to use this method in simulations of stratified compressible convection,
one need only adapt the BVP equations to the appropriate equations of hydrostatic equilibrium,
$\grad P = -\rho \bm{g}$, and thermal equilibrium, $\Div{F_{\text{cond}} + F_{\text{conv}}} = \text{sources}$,
for the problem at hand.  In compressible convection, where the density is allowed to change,
one must also use the knowledge of stellar structure codes and add an equation of mass conservation
in order to ensure that the BVP does not spuriously add or remove mass from the system.

This method can be extended to other boundary conditions, as well.  To solve for fixed temperature
boundary conditions, the difficulty is in finding the amount of flux through the system -- but
this can be done by using the ratios $f_{\text{conv}}$ and $f_{\text{cond}}$.  In the case of
fixed flux boundary conditions, there is degeneracy in the temperature solution which can come
out of the BVP, but in using knowledge about the system -- such as the initial symmetry of the
RB state around $T_1 = 0$, the final solution can be pegged onto the proper profile.
